### 2017-04-22 sections/implementation.tex
Update on Overleaf.

The objective of the compression algorithm is to reduce amount of data representing FEM results and also the ability to reconstruct original data from its smaller representation. This saves storage capacity and also accelerates data transfer between computers as the analysis itself and the post-processing of results is usually done on different work stations.

Compression method can be lossy or lossless based on quality of data reconstructed of its compressed representation. Lossless methods are able to fully reconstruct original data. Lossy methods on the other hand produce only approximations of original data.

@@

SVD decomposition is applied on matrices. The first thing to do is therefore to assemble an input matrix. Results from the Finite element method are discrete values calculated in nodes or integration points. There are usually multiple fields in result data (e.g. displacements, stress, strain etc.) and each field can have multiple components (e.g. displacements can have $X$, $Y$ and $Z$ component). As non-linear analyses usually lead to multiple computation steps there are multiple sets of data per each field component. All data corresponding to a single field component form the input matrix $\mtrx{A}$ to the compression algorithm. Only single field component is considered to fill the input matrix $\mtrx{A}$. Results for different data components (even for the same field) can be independent from each other. There can also be order-of-magnitude difference between the components and the compression algorithm could potentially clear the important information in the weaker component. Another reason against merging of components is computational complexity. The smaller the matrix $\mtrx{A}$ is, the faster the decomposition algorithm performs.

@@

From the definition of SVD in (\ref{eq:svd-def}) and from the properties of SVD presented in section \ref{sec:math} follows the fact that a matrix can be represented in the form of its SVD components as a sum of rank-1 matrices in the form

@@

\noindent
where $S_{i}$ is $i$-th singular value, $\mtrx{U_{i}}$ and $\mtrx{V_{i}}$ are corresponding singular vectors, and $n$ is the rank of matrix~$\mtrx{A}$. Considering the fact that singular values are ordered $S_{1} \geq S_{2} \geq S_{3} \geq ... \geq S_{n}$ the above formula implies that the first term would have the highest contribution and last term would have the lowest contribution to matrix~$\mtrx{A}$. Therefore, if we take only first $r$ members of the above summation we get an approximation matrix

@@

The quality of approximation depends on the magnitude of the singular values omitted from the approximation formula, namely $S_{r+1} ...  S_{n}$. The compression algorithm is based on assumption that the first singular value is order-of-magnitude higher than singular values at the end of the decomposition sequence. In special cases, when $r=n$, or $S_{i}=0$ for all $i \geq r$ the omitted singular values do not contribute to the sum and the compression is therefore lossless. In other cases, approximation error has to be calculated and taken into account to avoid loss of important details in data.

@@

Let us assume that the matrix is not empty and has full-rank. Then from the formula above follows that if $r$ equals to the rank of matrix $\mtrx{A}$ the compression ratio is always higher than one. In other words the memory consumption of stored decomposition is bigger than the size of the original matrix. To make the compression algorithm applicable the parameter $r$ must conform to the condition

$$r<\frac{m n}{m+n+1}.$$

@@

\begin{equation}
r=\ceil*{c \times \frac{m n}{m+n+1}},
\end{equation}

@@

\paragraph{Approximation error}
In the usual case the most important parameter to take into account is the approximation error. Algorithm is trying to minimize the compression ratio while at the same time ensuring that predefined limit of approximation error is not exceeded. To quantify the error the Mean squared error ($MSE$) metric is used. To be able to compare different data-sets having different scales the normalized version of $MSE$ is used - Normalized root-mean-square deviation ($NRMSD$). Both metrics are defined in section \ref{sec:error}.

@@

Then $NRMSD$ can be used as a quality metric for the compression algorithm because normalization makes it usable for different datasets. Calculation of rank of the approximation matrix is depicted as pseudo-code in algorithm \ref{alg:rank-calculation}. Algorithm uses the following inequality to test whether the desired rank has been reached

@@

\begin{itemize}
\item The algorithm is most efficient in cases where one dimension of the input matrix is very low compared to the other. However, this is almost always the case when compressing results from FEM -- number of time steps seldom exceeds hundreds.
\item Moreover, time steps can be split into smaller ranges and the algorithm can be applied on each range separately. This will improve performance and can also increase quality of compression if the key time steps on the range boundaries are carefully selected.
\item Randomized SVD algorithm has the same order of algorithmic complexity but yet can help to significantly reduce execution time.
\end{itemize}

>>>>

The objective of a compression algorithm is to reduce amount of data representing FEM results and also the ability to reconstruct original data from its smaller representation. This saves storage capacity and also accelerates data transfer between computers as the analysis itself and the post-processing of results is usually done on different work stations.

Compression method can be lossy or lossless based on quality of data reconstructed from its compressed representation. Lossless methods are able to fully reconstruct original data. Lossy methods on the other hand produce only approximations of original data.

@@

SVD decomposition is applied on matrices. The first thing to do is therefore to assemble an input matrix. Results from the Finite element method are discrete values calculated in nodes or integration points. There are usually multiple fields in result data (e.g. displacements, stress, strain etc.) and each field can have multiple components (e.g. displacements can have $X$, $Y$ and $Z$ component). As non-linear analyses usually lead to multiple computation steps there are multiple sets of data per each field component. All data corresponding to a single field component form the input matrix $\mtrx{A}$ to the compression algorithm. Only single field component is considered to fill the input matrix $\mtrx{A}$. Results for different data components (even for the same field) can be independent from each other. There can also be order-of-magnitude difference between the components and the compression algorithm could potentially clear the important information in the weaker component. Another reason against merging of components is computational complexity. The smaller the matrix $\mtrx{A}$ is the faster the decomposition algorithm performs.

@@

From the definition of SVD in (\ref{eq:svd-def}) and from the properties of SVD presented in section \ref{sec:math} follows the fact that a matrix can be represented in the form of its SVD components as a sum of rank-1 matrices

@@

\noindent
where $S_{i}$ is the $i$-th singular value, $\mtrx{U_{i}}$ and $\mtrx{V_{i}}$ are corresponding singular vectors, and $n$ is the rank of matrix~$\mtrx{A}$. Considering the fact that singular values are ordered $S_{1} \geq S_{2} \geq S_{3} \geq ... \geq S_{n}$, the above formula implies that the first term of the sum would have the highest contribution and the last term would have the lowest contribution to matrix~$\mtrx{A}$. Therefore, if we take only first $r$ members of the above summation we get an approximation matrix

@@

The quality of approximation depends on the magnitude of the singular values omitted from the approximation formula, namely $S_{r+1} ...  S_{n}$. The compression algorithm is based on an assumption that the first singular value is order-of-magnitude higher than singular values at the end of the decomposition sequence. In special cases, when $r=n$ or $S_{i}=0$ for all $i \geq r$, the omitted singular values do not contribute to the sum and the compression is therefore lossless. In other cases, approximation error has to be calculated and taken into account to avoid loss of important details in data.

@@

Let us assume that the matrix is not empty and is full rank. Then from the formula (\ref{eq:cr-def}) follows that if $r$ equals to the rank of matrix $\mtrx{A}$ the compression ratio is always higher than one. In other words the memory consumption of stored decomposition is bigger than the size of the original matrix. To make the compression algorithm applicable the parameter $r$ must conform to the condition

\begin{equation}
r<\frac{m n}{m+n+1}.
\label{eq:r-ineq}
\end{equation}

@@

\begin{equation}
r=\ceil*{c \times \frac{m n}{m+n+1}},
\label{eq:rank-from-comp-ratio}
\end{equation}

@@

\paragraph{Approximation error}
In the usual case the most important parameter to take into account is the approximation error. Algorithm is trying to minimize the compression ratio while at the same time ensuring that predefined approximation error threshold is not exceeded. To quantify the error the Mean squared error ($MSE$) metric is used. To be able to compare different data-sets having different scales the normalized version of $MSE$ is used - Normalized root-mean-square deviation ($NRMSD$). Both metrics are defined in section \ref{sec:error}.

@@

Then $NRMSD$ can be used as a quality metric for the compression algorithm because normalization makes it usable for different datasets. Calculation of rank of the approximation matrix is depicted as pseudo-code in Algorithm \ref{alg:rank-calculation}. Algorithm uses the following inequality to test whether the desired rank has been reached

@@

\begin{itemize}
\item The algorithm is most efficient in cases where one dimension of the input matrix is very low compared to the other. However, this is almost always the case when compressing results from FEM -- number of time steps seldom exceeds hundreds.
\item Moreover, time steps can be split into smaller ranges and the algorithm can be applied on each range separately. This will improve performance and can also increase quality of compression if the key time steps on the range boundaries are carefully selected.
\item Randomized SVD algorithm has the same order of algorithmic complexity but yet can significantly reduce execution time. (The implementation of the particular randomized SVD algorithm that was used is described in \cite{Halko2011}.)
\end{itemize}

### 2017-04-22 sections/introduction.tex
Update on Overleaf.

The effort to reduce size of resulting data from complex finite element analyses to accelerate (or even make possible) post-processing using common personal computer is not new.

% TODO: State of the art. Add references to work in the field of FEM results post-processing.

In \cite{Benes2016} is presented one possible direction for research in this area -- the replacement of discrete data produced by finite element solver by simple continuous functions. These approximation functions can be then described by a small number of parameters. The focus is on geometry. The main goal is to find the areas in domain where the output discrete function has predictable development and can be easily replaced by e.g. linear continuous function. Although this approach has great results for some class of functions it is not applicable to problem in general. For some special cases, such as functions with discontinuities, the method has poor results because approximation error is too high and -- what is more important -- it cannot be controlled in advance.

In this study it was decided to apply purely algebraic approach. Various methods were considered, e.g. discrete wavelet transform \cite{Lui2001} and discrete cosine transform \cite{Watson1994} for their successful use in image compression. The reason why the SVD method was finally chosen is that it is well suited for all kind of data and the implementation is straightforward, considering that the results from the FEM analyses can be viewed as a series of arbitrary rectangular matrices.

% TODO: Add references to related work and related usage of SVD for compression

% TODO: Description of all terms that are used in text?

>>>>

The effort to reduce size of resulting data from complex finite element analyses to accelerate (or even make possible) post-processing using common personal computer is not new. In \cite{Benes2016} is presented one possible direction for research in this area -- the replacement of discrete data produced by finite element solver by continuous functions. These approximation functions can be then described by a small number of parameters. The focus is on geometry. The main goal is to find the areas in domain where the output discrete function has predictable development and can be easily replaced by a simple continuous function, e.g. linear function. Although this approach has great results for some class of functions it is not applicable to problem in general. For some special cases, such as functions with discontinuities, the method has poor results because approximation error is too high and -- what is more important -- it cannot be determined in advance.

In this study it was decided to apply purely algebraic approach. Various methods were considered, e.g. discrete wavelet transform \cite{Lui2001} and discrete cosine transform \cite{Watson1994}, for their successful use in image compression. The reason why the SVD method was finally chosen is that it is well suited for all kinds of data and the implementation is straightforward, considering that the results from the FEM analyses can be viewed as a series of arbitrary rectangular matrices.

% State of the art. Add references to work in the field of FEM results post-processing.
% Add references to related work and related usage of SVD for compression
% Description of all terms that are used in text?

### 2017-04-22 sections/math.tex
Update on Overleaf.

\noindent
where $\mtrx{U^\mathsf{T}U} = \mtrx{I}$, $\mtrx{V^\mathsf{T}V} = \mtrx{I}$; the columns of $\mtrx{U}$ are orthonormal eigenvectors of $\mtrx{AA^\mathsf{T}}$, the columns of $\mtrx{V}$ are orthonormal eigenvectors of $\mtrx{A^\mathsf{T}A}$, and $\mtrx{S}$ (sometimes referred to as $\mtrx{\Sigma}$) is a diagonal matrix containing singular values in descending order which are at the same time the square roots of eigenvalues of $\mtrx{U}$ or $\mtrx{V}$.

SVD can be seen as a method for transforming correlated variables into a set of uncorrelated ones. At the same time, SVD is a method for identifying and ordering the dimensions along which data points show the most variation. Once we have identified where the most variation is, it is possible to find the best approximation of the original data points using fewer dimensions. Hence, SVD can be seen as a method for data reduction/compression.

@@

\noindent
where $X_{min}$ and $X_{max}$ are elements of input matrix $\mtrx{A}$ with minimum and maximum value, respectively. Being able to compare datasets with different scales this is the main parameter that is used to control quality of compression in the algorithm presented in this paper.

\item \textbf{Peak signal to noise ratio}

>>>>

\noindent
where $\mtrx{U^\mathsf{T}U} = \mtrx{I}$, $\mtrx{V^\mathsf{T}V} = \mtrx{I}$; the columns of $\mtrx{U}$ are orthonormal eigenvectors of $\mtrx{AA^\mathsf{T}}$, the columns of $\mtrx{V}$ are orthonormal eigenvectors of $\mtrx{A^\mathsf{T}A}$, and $\mtrx{S}$ (sometimes referred to as $\mtrx{\Sigma}$) is a diagonal matrix containing singular values in descending order, which are at the same time the square roots of eigenvalues of $\mtrx{U}$ or $\mtrx{V}$.

SVD can be seen as a method for transforming correlated variables into a set of uncorrelated ones. At the same time, SVD is a method for identifying and ordering the dimensions along which data points show the most variation. Once it has been identified where the most variation is, it is possible to find the best approximation of the original data points using fewer dimensions. Hence, SVD can be seen as a method for data reduction/compression.

@@

\noindent
where $X_{min}$ and $X_{max}$ are elements of input matrix $\mtrx{A}$ with minimum and maximum value, respectively. This error metric is able to measure and compare errors in datasets with different scales. Therefore, it is the main parameter that is used to control the quality of compression in the algorithm presented in this paper.

\item \textbf{Peak Signal to Noise Ratio}
